---
title: "FitbitAnalysis"
author: "Bhagyashri Mane"
date: "2025-03-06"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Step 1: Load the packages.

```{r}
install.packages("tidyverse")
install.packages("skimr")
install.packages("cowplot")     #used for combining multiple ggplot2 plots into a single figure
install.packages("plotly")
install.packages("tidyr")   #reshape and organize data into a tidy format 
install.packages("janitor")

library(tidyverse)        #wrangle data
library(plotly)           #interactive visualizations
library(dplyr)            #clean data
library(lubridate)        #wrangle date attributes
library(skimr)            #get summary data
library(ggplot2)          #visualize data
library(cowplot)          #grid the plot
library(readr)            #save csv 
library(plotly)           #pie chart
library(janitor)          #data cleaning
```

# Step 2: Prepare the data and combine them in one data frame.

```{r}
setwd("/cloud/project")  #set working directory

daily_activity <- read.csv("dailyActivity_merged.csv")
sleep_day <- read.csv("sleepDay_merged.csv")
weight <- read.csv("weightLogInfo_merged.csv")

hourly_steps <- read.csv("hourlySteps_merged.csv")
hourly_intensities <-read.csv("hourlyIntensities_merged.csv")
hourly_calories <- read.csv("hourlyCalories_merged.csv")

# heartrate <- read.csv("heartrate_seconds_merged.csv")

```

## take a look at all dataframes

```{r}
head(daily_activity) # preview of the dataframe
colnames(daily_activity) # gives column names
glimpse(daily_activity) #summary of dataframe
```

```{r}
head(sleep_day)
colnames(sleep_day)
glimpse(sleep_day)
```

```{r}
head(weight)
colnames(weight)
glimpse(weight)
```

```{r}
head(hourly_steps)
colnames(hourly_steps)
glimpse(hourly_steps)
```

```{r}
head(hourly_intensities)
colnames(hourly_intensities)
glimpse(hourly_intensities)
```

```{r}
head(hourly_calories)
colnames(hourly_calories)
glimpse(hourly_calories)
```

```{r}
# head(heartrate)
# colnames(heartrate)
# glimpse(heartrate)
```

## Check for NA

```{r}
sum(is.na(daily_activity))
sum(is.na(sleep_day))
sum(is.na(weight))
sum(is.na(hourly_steps))
sum(is.na(hourly_intensities))
sum(is.na(hourly_calories))
# sum(is.na(heartrate))
## We will leave the NA. The NA belong to "Fat" data of different dates.
```

## check for duplicates and remove duplicates

```{r}
cat("Number of duplicate rows in daily_activity before removal:", nrow(daily_activity[duplicated(daily_activity),]), "\n")
daily_activity <- daily_activity[!duplicated(daily_activity),]
cat("Number of duplicate rows in daily_activity after removal:", nrow(daily_activity[duplicated(daily_activity),]), "\n") 
# Confirm duplicates removed
sum(duplicated(daily_activity))

cat("Number of duplicate rows in sleep_day before removal:", nrow(sleep_day[duplicated(sleep_day),]), "\n")
sleep_day <- sleep_day[!duplicated(sleep_day),]
cat("Number of duplicate rows in sleep_day after removal:", nrow(sleep_day[duplicated(sleep_day),]), "\n") # Confirm duplicates removed
sum(duplicated(sleep_day))

cat("Number of duplicate rows in weight before removal:", nrow(weight[duplicated(weight),]), "\n")
weight <- weight[!duplicated(weight),]
cat("Number of duplicate rows in weight after removal:", nrow(weight[duplicated(weight),]), "\n") # Confirm duplicates removed
sum(duplicated(weight))

cat("Number of duplicate rows in hourly_steps before removal:", nrow(hourly_steps[duplicated(hourly_steps),]), "\n")
hourly_steps <- hourly_steps[!duplicated(hourly_steps),]
cat("Number of duplicate rows in hourly_steps after removal:", nrow(hourly_steps[duplicated(hourly_steps),]), "\n") # Confirm duplicates removed
sum(duplicated(hourly_steps))

cat("Number of duplicate rows in hourly_intensities before removal:", nrow(hourly_intensities[duplicated(hourly_intensities),]), "\n")
hourly_intensities <- hourly_intensities[!duplicated(hourly_intensities),]
cat("Number of duplicate rows in hourly_intensities after removal:", nrow(hourly_intensities[duplicated(hourly_intensities),]), "\n") # Confirm duplicates removed
sum(duplicated(hourly_intensities))

cat("Number of duplicate rows in hourly_calories before removal:", nrow(hourly_calories[duplicated(hourly_calories),]), "\n")
hourly_calories <- hourly_calories[!duplicated(hourly_calories),]
cat("Number of duplicate rows in hourly_calories after removal:", nrow(hourly_calories[duplicated(hourly_calories),]), "\n") # Confirm duplicates removed
sum(duplicated(hourly_calories))

# cat("Number of duplicate rows in heartrate before removal:", nrow(heartrate[duplicated(dailyheartrate_activity),]), "\n")
# heartrate <- heartrate[!duplicated(heartrate),]
# cat("Number of duplicate rows in heartrate after removal:", nrow(heartrate[duplicated(heartrate),]), "\n") # Confirm duplicates removed
# sum(duplicated(heartrate))

```

## Convert character columns to date

```{r}
daily_activity$ActivityDate <- as.Date(daily_activity$ActivityDate, format = "%m/%d/%Y")
sleep_day$SleepDay <- as.Date(sleep_day$SleepDay, format = "%m/%d/%Y")
weight$Date <- as.Date(weight$Date, format = "%m/%d/%Y")

hourly_steps$ActivityHour <- as.POSIXct(hourly_steps$ActivityHour, format = "%m/%d/%Y %I:%M:%S %p") # Convert ActivityHour to POSIXct (24-hour format )
hourly_steps$Date <- as.Date(format(hourly_steps$ActivityHour, format = "%m/%d/%Y")) # Extract Date
hourly_steps$Hour <- format(hourly_steps$ActivityHour, format = "%H")  # Character # Extract the hour
hourly_steps$Hour <- as.numeric(hourly_steps$Hour)            # Convert to numeric (0-23)
write_csv(hourly_steps, "hourly_steps.csv")                 # Check result

hourly_intensities$ActivityHour <- as.POSIXct(hourly_intensities$ActivityHour, format = "%m/%d/%Y %I:%M:%S %p")
hourly_intensities$Date <- as.Date(format(hourly_intensities$ActivityHour, format = "%m/%d/%Y"))
hourly_intensities$Hour <- format(hourly_intensities$ActivityHour, format = "%H")  # Character
hourly_intensities$Hour <- as.numeric(hourly_intensities$Hour)  # Convert to numeric (0-23)
write_csv(hourly_intensities, "hourly_intensities.csv")

hourly_calories$ActivityHour <- as.POSIXct(hourly_calories$ActivityHour, format = "%m/%d/%Y %I:%M:%S %p")
hourly_calories$Date <- as.Date(format(hourly_calories$ActivityHour, format = "%m/%d/%Y"))
hourly_calories$Hour <- format(hourly_calories$ActivityHour, format = "%H")  # Character
hourly_calories$Hour <- as.numeric(hourly_calories$Hour)  # Convert to numeric (0-23)
write_csv(hourly_calories, "hourly_calories.csv")

# heartrate$Time <- as.POSIXct(heartrate$Time, format = "%m/%d/%Y %H:%M:%S")

```

## Add a new column for the weekdays

```{r}
daily_activity <- daily_activity %>% mutate( weekday = weekdays(as.Date(ActivityDate, "%m/%d/%Y")))

```

## Cleaning column names

```{r}
daily_activity <- daily_activity %>% clean_names()
sleep_day <- sleep_day %>% clean_names()
weight <- weight %>% clean_names()

hourly_steps <- hourly_steps %>% clean_names()
hourly_intensities <- hourly_intensities %>% clean_names()
hourly_calories <- hourly_calories %>% clean_names()

# heartrate <- heartrate %>% clean_names()
```

## View updated column names

```{r}
colnames(daily_activity)

colnames(sleep_day)

colnames(weight)

colnames(hourly_steps)

colnames(hourly_intensities)

colnames(hourly_calories)

# colnames(heartrate)

```

## merge daily_activity, sleep and weight data

```{r}
merged1 <- merge(daily_activity,sleep_day,
                  by.x = c("id", "activity_date"), 
                  by.y = c("id", "sleep_day"),
                  all=TRUE)
merged_data <- merge(merged1, weight, 
                      by.x = c("id", "activity_date"), 
                      by.y = c("id", "date"),
                      all=TRUE)

#save csv for future use
write_csv(merged_data, "merged_data.csv")

# Check head, NA, duplicates, distinct id in merged data.
head(merged_data)
glimpse(merged_data)
sum(is.na(merged_data))
sum(duplicated(merged_data))
n_distinct(merged_data$id)

```

## merge hourly steps, intensity and calories data

```{r}
 hourly_merge_one <- merge(hourly_steps,hourly_intensities,
                      by = c("id", "date", "hour"), 
                    all=TRUE)
 hourly_merge <- merge(hourly_merge_one, hourly_calories, 
                       by = c("id", "date", "hour"), 
                       all=TRUE)
#save csv for future use
 write_csv(hourly_merge, "hourly_merge.csv")
# Check head, NA, duplicates, distinct id in merged data.
head(hourly_merge)
glimpse(hourly_merge)
sum(is.na(hourly_merge))
sum(duplicated(hourly_merge))
n_distinct(hourly_merge$id)
```

```{r}
hourly_merge <- read.csv("hourly_merge.csv") # because i have manually added the merge file
hourly_merge <- hourly_merge %>% clean_names()
colnames(hourly_merge)
```

## Order from Monday to Sunday for plot later

```{r}
merged_data$weekday <- factor(merged_data$weekday, levels= c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")) #sequence
merged_data[order(merged_data$weekday), ]                   #order by weekday
head(merged_data)
colnames(merged_data)
```

## Summarize key metrics

```{r}

daily_activity %>%
  summarise(participants = n_distinct(id))
sleep_day %>%
  summarise(participants = n_distinct(id)) # we got 24
weight %>%
  summarise(participants = n_distinct(id)) # we got 8

hourly_steps %>%
  summarise(participants = n_distinct(id))
hourly_intensities %>%
  summarise(participants = n_distinct(id))
hourly_calories %>%
  summarise(participants = n_distinct(id))

#heartrate %>%
#  summarise(participants = n_distinct(id))

```

# Step 3: Examine the dataset and check if all 30 users are unique.

```{r}
## Check to see if all users are unique.We supposed to have 30 users or 30 IDs. So We have 3 extra from daily activity, 6 less from the sleep day table, and 22 less from the weight table. 
n_distinct(daily_activity$id)
n_distinct(sleep_day$id)
n_distinct(weight$id)

## Since weight table only has 8 users enter their information. Let's take a look at how they enter the information. 5 users are manually reporting the weight and 3 uers are reporting it with a connected device - wifi connected scale. 
weight %>% 
  filter(is_manual_report == "True") %>% 
  group_by(id) %>% 
  summarise("Manual Weight Report"=n()) %>%
  distinct()

weight %>% 
  filter(is_manual_report == "False") %>% 
  group_by(id) %>% 
  summarise("Manual Weight Report"=n()) %>%
  distinct()


## When are users most active in recording their data. We noticed users track their data more from Tuesday to Thursday and we have more of those days' data than other days.
most_data_recorder_on <- merged_data %>%
  group_by(weekday) %>%
  summarise(record_count = n(), .groups = "drop")
glimpse(most_data_recorder_on)

most_active_days <- ggplot(data=most_data_recorder_on, aes(x=weekday, y=record_count))+
  geom_bar(stat = "identity",fill="#CD69C9")+
  labs(title="Data distribution over the week")+
  theme_minimal()
ggsave("Data_distribution_over_the_week.png", plot = most_active_days, width = 6, height = 3, dpi = 200) #plot1
```

# Step 4: Weekly summary

```{r}
# Weekly analysis of Steps 
# calculate the avg of all over steps for each weekday here group only by weekday
avg_weekday_steps <- merged_data %>%
                      group_by(weekday) %>%
                      summarise(avg_weekly_steps = mean(total_steps, na.rm = TRUE), .groups = "drop")
glimpse(avg_weekday_steps)

Average_Weekday_Steps <-  ggplot(data=avg_weekday_steps, aes(x=weekday, y=avg_weekly_steps, fill=weekday))+ 
                          geom_bar(stat="identity")+
                          geom_text(aes(label = round(avg_weekly_steps, 1)), vjust = -0.5, size = 3) +
                          labs(title="Average Steps Per Day", x= "Day of the week" ,y="Average Steps")+
                          theme_minimal()
ggsave("Average_Weekday_Steps.png", plot = Average_Weekday_Steps, width = 8, height = 6, dpi = 200) #plot2

# Weekly analysis of Distance
filtered_data <- merged_data %>% filter(!is.na(total_distance))

avg_weekday_distance <- merged_data %>%
                        group_by(weekday) %>%
                        summarise(avg_weekly_distance = mean(total_distance, na.rm = TRUE), .groups = "drop")
glimpse(avg_weekday_distance)


Average_Weekday_Distance <-  ggplot(data=avg_weekday_distance, aes(x=weekday, y=avg_weekly_distance, fill=weekday))+ 
                          geom_bar(stat="identity")+
                          geom_text(aes(label = round(avg_weekly_distance, 1)), vjust = -0.5, size = 3) +
                          labs(title="Average Distance Per Day",x="Day of the week", y="Average Distance")+
                          theme_minimal()

ggsave("Average_Weekday_Distance.png", plot = Average_Weekday_Distance, width = 8, height = 6, dpi = 200) #plot3

# Similar we can plot for calories and Intensity

# Weekly analysis of sedentary_minutes

avg_weekday_sedmin <-   merged_data %>%
                        group_by(weekday) %>%
                        summarise(avg_weekly_sedmin = mean(sedentary_minutes, na.rm = TRUE), .groups = "drop")
glimpse(avg_weekday_sedmin)


plot3 <- ggplot(data=avg_weekday_sedmin, aes(x=weekday, y=avg_weekly_sedmin, fill=weekday))+ 
         geom_bar(stat="identity")+
         geom_text(aes(label = round(avg_weekly_sedmin, 1)), vjust = -0.5, size = 3) +
         labs(title="Average Sedentary Minutes Per Day",x="Day of the week", y="Sedentary Minutes")+
         theme_minimal()
ggsave("Average_Weekday_sedmin.png", plot = plot3, width = 8, height = 6, dpi = 200) #plot4

# Weekly analysis of time asleep

avg_weekday_time_asleep <-   merged_data %>%
                        group_by(weekday) %>%
                        summarise(avg_weekly_time = mean(total_minutes_asleep, na.rm = TRUE), .groups = "drop")
glimpse(avg_weekday_time_asleep)


plot4 <- ggplot(data=avg_weekday_time_asleep, aes(x=weekday, y=avg_weekly_time, fill=weekday))+ 
         geom_bar(stat="identity")+
         geom_text(aes(label = round(avg_weekly_time, 1)), vjust = -0.5, size = 3) +
         labs(title="Average Asleep Time Per Day",x="Day of the week", y="Time Asleep")+
         theme_minimal()
ggsave("Average_Weekday_Sleep_Time.png", plot = plot4, width = 8, height = 6, dpi = 200) #plot5


```

# Step 5: hourly summary

```{r}
avg_hourly_steps <-  hourly_merge %>%
                    filter(step_total > 0) %>% 
                    group_by(hour) %>%
                    summarise(avg_steps = mean(step_total, na.rm = TRUE), .groups = "drop")
glimpse(avg_hourly_steps)


plot6 <-  ggplot(data=avg_hourly_steps, aes(x=hour, y=avg_steps, fill=avg_steps ))+
  geom_bar(stat="identity")+
  geom_text(aes(label = round(avg_steps)), vjust = -0.5, size = 3)+
  scale_fill_gradientn(colors = c("red", "orange", "yellow", "green"))+
  scale_x_continuous(breaks = 0:23) +  # Ensure all hours (0 to 23) are displayed
  labs(title="Hourly Steps", x="Hour of the Day", y="Average Steps") +
  theme_minimal()
ggsave("Average_Hourly_Steps.png", plot = plot6, width = 10, height = 5, dpi = 200) #plot6

```

# Step 6: Statistics summary mean, median, min, max for all 3 tables + merged data

```{r}

daily_activity %>%
 dplyr::select(total_steps,
         total_distance,
         very_active_minutes,
         fairly_active_minutes,
         lightly_active_minutes,
         sedentary_minutes,
         calories) %>%
         na.omit() %>% #remove missing values before summarizing
         summary()


sleep_day %>%  
  dplyr::select(total_sleep_records,
          total_minutes_asleep,
          total_time_in_bed) %>%
          na.omit() %>% #remove missing values before summarizing
          summary()


weight %>% 
  dplyr::select(weight_pounds, bmi) %>%
  na.omit() %>%
  summary()


merged_data %>%
  dplyr::select(
         total_steps,
         total_distance,
         very_active_minutes,
         fairly_active_minutes,
         lightly_active_minutes,
         sedentary_minutes,
         calories,
         total_minutes_asleep,
         total_time_in_bed,
         weight_kg,
         bmi
         ) %>%
  na.omit() %>%
  summary()


hourly_merge %>%
  dplyr::select(
                hour,
                step_total,
                calories,
                total_intensity,
                )%>%
  na.omit() %>%
  summary()
  
```


# Step 7: analysis on active minutes, calorie, total steps. 
#### The American Heart Association and World Health Organization recommend at least 150 minutes of moderate-intensity activity or 75 minutes of vigorous activity,or a combination of both, each week. That means it needs an daily goal of 21.4 minutes of FairlyActiveMinutes or 10.7 minutes of VeryActiveMinutes

```{r}
# Active users activity distribution in percentage
active_users <- merged_data %>%
  filter(fairly_active_minutes >= 21.4 | very_active_minutes>=10.7) %>% 
  group_by(id) %>% 
  count(id) 
glimpse(active_users)

total_minutes <- sum(merged_data$sedentary_minutes, 
                     merged_data$very_active_minutes, 
                     merged_data$fairly_active_minutes, 
                     merged_data$lightly_active_minutes)

sedentary_percentage <- sum(merged_data$sedentary_minutes)/total_minutes*100
lightly_percentage <- sum(merged_data$lightly_active_minutes)/total_minutes*100
fairly_percentage <- sum(merged_data$fairly_active_minutes)/total_minutes*100
active_percentage <- sum(merged_data$very_active_minutes)/total_minutes*100

## Pie charts
percentage <- data.frame(
  level=c("Sedentary", "Lightly", "Fairly", "Very Active"),
  minutes=c(sedentary_percentage,lightly_percentage,fairly_percentage,active_percentage)
  )


plot_ly(percentage, labels = ~level, values = ~minutes, type = 'pie',textposition = 'outside',textinfo = 'label+percent',
        marker = list(colors = c("#FF4500", "#1E90FF", "#32CD32", "#FFD700"))) %>%
          layout(title = 'Activity Minutes Percentage',
          xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
          yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE)
          )

# How active are the users
# library(tidyr)

active_minute <- merged_data %>% 
                pivot_longer(cols = ends_with("minutes"), 
               names_to = "intensity", 
               values_to = "active_minutes")%>% 
  select(intensity, active_minutes) 

  
plot8 <- ggplot(data=active_minute, aes(x=intensity, y=active_minutes))+
        geom_boxplot(aes(fill=intensity))+
        scale_x_discrete(limits=c("sedentary_minutes","lightly_active_minutes","fairly_active_minutes","very_active_minutes"))+
        ylab("Minutes")+
        theme_minimal()
ggsave("boxplot.png", plot = plot8, width = 8, height = 6, dpi = 300) #plot8

## Interesting find here that some user who are sedentary, takes minimal step, but still able to burn over 1500 to 2500 calories
plot9 <-
        ggplot(data=merged_data, aes(x=total_steps, y = calories, color=sedentary_minutes))+ 
  geom_point()+ 
  stat_smooth(method=lm)+
  scale_color_gradient(low="steelblue", high="orange")+
  labs(title="The Relationship Between Calories and Total Steps", x="Total Steps", y="Calories Burned")+
  theme_minimal()
ggsave("calories_vs_total_steps.png", plot = plot9, width = 8, height = 6, dpi = 300) #plot9

```

# Step 8: This high volume of moderate-to-vigorous physical activity is achieved by a very small proportion of the population. Let's take a look at this.

```{r}
# active_minutes_vs_calories

active_minutes_vs_calories <- ggplot(data = merged_data) + 
  geom_point(mapping=aes(x=calories, y=fairly_active_minutes), color = "maroon", alpha = 1/3) +
  geom_smooth(method = loess,formula =y ~ x, mapping=aes(x=calories, y=fairly_active_minutes, color=fairly_active_minutes), color = "maroon", se = FALSE) +
  geom_point(mapping=aes(x=calories, y=very_active_minutes), color = "forestgreen", alpha = 1/3) +
  geom_smooth(method = loess,formula =y ~ x,mapping=aes(x=calories, y=very_active_minutes, color=very_active_minutes), color = "forestgreen", se = FALSE) +
  geom_point(mapping=aes(x=calories, y=lightly_active_minutes), color = "orange", alpha = 1/3) +
  geom_smooth(method = loess,formula =y ~ x,mapping=aes(x=calories, y=lightly_active_minutes, color=lightly_active_minutes), color = "orange", se = FALSE) +
  geom_point(mapping=aes(x=calories, y=sedentary_minutes), color = "steelblue", alpha = 1/3) +
  geom_smooth(method = loess,formula =y ~ x,mapping=aes(x=calories, y=sedentary_minutes, color=sedentary_minutes), color = "steelblue", se = FALSE) +
  annotate("text", x=4800, y=160, label="Very Active", color="black", size=3)+
  annotate("text", x=4800, y=0, label="Fairly Active", color="black", size=3)+
  annotate("text", x=4800, y=500, label="Sedentary", color="black", size=3)+
  annotate("text", x=4800, y=250, label="Lightly  Active", color="black", size=3)+
  labs(x = "Calories", y = "Active Minutes", title="Active Minutes vs Calories")
ggsave("active_minutes_vs_calories.png", plot = active_minutes_vs_calories, width = 14, height = 10, dpi = 300) #plot10

# active_minutes_vs_steps
active_minutes_vs_steps <- ggplot(data = merged_data) + 
  geom_point(mapping=aes(x=total_steps, y=fairly_active_minutes), color = "maroon", alpha = 1/3) +
  geom_smooth(method = loess,formula =y ~ x, mapping=aes(x=total_steps, y=fairly_active_minutes, color=fairly_active_minutes), color = "maroon", se = FALSE) +
  
  geom_point(mapping=aes(x=total_steps, y=very_active_minutes), color = "forestgreen", alpha = 1/3) +
  geom_smooth(method = loess,formula =y ~ x,mapping=aes(x=total_steps, y=very_active_minutes, color=very_active_minutes), color = "forestgreen", se = FALSE) +
  
  geom_point(mapping=aes(x=total_steps, y=lightly_active_minutes), color = "orange", alpha = 1/3) +
  geom_smooth(method = loess,formula =y ~ x,mapping=aes(x=total_steps, y=lightly_active_minutes, color=lightly_active_minutes), color = "orange", se = FALSE) +
  
   geom_point(mapping=aes(x=total_steps, y=sedentary_minutes), color = "steelblue", alpha = 1/3) +
  geom_smooth(method = loess,formula =y ~ x,mapping=aes(x=total_steps, y=sedentary_minutes, color=sedentary_minutes), color = "steelblue", se = FALSE) +
  
  annotate("text", x=35000, y=150, label="Very Active", color="black", size=3)+
  annotate("text", x=35000, y=50, label="Fairly Active", color="black", size=3)+
  annotate("text", x=35000, y=1350, label="Sedentary", color="black", size=3)+
  annotate("text", x=35000, y=380, label="Lightly  Active", color="black", size=3)+
  labs(x = "Total Steps", y = "Active Minutes", title="Active Minutes vs Steps")
ggsave("active_minutes_vs_steps.png", plot = active_minutes_vs_steps, width = 14, height = 10, dpi = 300) #plot11


# active_minutes_vs_distance
active_minutes_vs_distance <- ggplot(data = merged_data) + 
  geom_point(mapping=aes(x=total_distance, y=fairly_active_minutes), color = "steelblue", alpha = 1/3) +
  geom_smooth(method = loess,formula =y ~ x, mapping=aes(x=total_distance, y=fairly_active_minutes, color=fairly_active_minutes), color = "steelblue", se = FALSE) +
  
  geom_point(mapping=aes(x=total_distance, y=very_active_minutes), color = "gold", alpha = 1/3) +
  geom_smooth(method = loess,formula =y ~ x,mapping=aes(x=total_distance, y=very_active_minutes, color=very_active_minutes), color = "gold", se = FALSE) +
  
  geom_point(mapping=aes(x=total_distance, y=lightly_active_minutes), color = "coral", alpha = 1/3) +
  geom_smooth(method = loess,formula =y ~ x,mapping=aes(x=total_distance, y=lightly_active_minutes, color=lightly_active_minutes), color = "coral", se = FALSE) +
  
   geom_point(mapping=aes(x=total_distance, y=sedentary_minutes), color = "forestgreen", alpha = 1/3) +
  geom_smooth(method = loess,formula =y ~ x,mapping=aes(x=total_distance, y=sedentary_minutes, color=sedentary_minutes), color = "forestgreen", se = FALSE) +
  
  scale_x_continuous(limits = c(0, 30))+
  
  annotate("text", x=28, y=150, label="Very Active", color="black", size=3)+
  annotate("text", x=28, y=50, label="Fairly Active", color="black", size=3)+
  annotate("text", x=28, y=1250, label="Sedentary", color="black", size=3)+
  annotate("text", x=28, y=280, label="Lightly  Active", color="black", size=3)+
  labs(x = "Distance", y = "Active Minutes", title = "Activity Minutes vs Distance")
ggsave("active_minutes_vs_distance.png", plot = active_minutes_vs_distance, width = 14, height = 10, dpi = 300) #plot12
```

# Step 9: Analysis on sleep

```{r}
## Sleep time in hours instead of minutes
sleep_day_in_hour <- sleep_day
sleep_day_in_hour$total_minutes_asleep <- sleep_day_in_hour$total_minutes_asleep/60
sleep_day_in_hour$total_time_in_bed <- sleep_day_in_hour$total_time_in_bed/60
head(sleep_day_in_hour)

## Check for any sleep outliers. # of times user sleep more than 10 hours or less than 1  
sum(sleep_day_in_hour$total_minutes_asleep > 9)
sum(sleep_day_in_hour$total_time_in_bed > 9)
sum(sleep_day_in_hour$total_minutes_asleep < 2)
sum(sleep_day_in_hour$total_time_in_bed < 2)

## According to article: https://blog.fitbit.com/sleep-study/#:~:text=The%20average%20Fitbit%20user%20is,is%20spent%20restless%20or%20awake.&text=People%20who%20sleep%205%20hours,the%20beginning%20of%20the%20night. 55 minutes are spend awake in bed before going to sleep. Let see how many users in our study is according to the FitBit data

awake_in_bed <- mutate(sleep_day, AwakeTime = total_time_in_bed - total_minutes_asleep)
awake_in_bed <- awake_in_bed %>% 
  filter(AwakeTime >= 55) %>% 
  group_by(id) %>% 
  arrange(AwakeTime, desc=TRUE) 
  
n_distinct(awake_in_bed$Id) #13 users spend more than 55 minutes in bed before falling alseep

## How many minutes an user sleep may not correlate well with how actively they are, but sedentary time account for about 80% of during the day

## Majority of the users sleep between 5 to 10 hours burns around 1500 to 4500 calories a day.
ggplot(data=merged_data, aes(x=total_minutes_asleep/60, y=calories, color=total_steps))+ 
  geom_point()+
  geom_smooth(method = 'loess', formula = 'y ~ x', color = '#005FCC') +
  scale_color_gradient(low="steelblue", high="orange")+
  labs(x = "Sleep Hours", y = "Calories", title = "Calories vs Sleep Hours")

## Majority of the users sleep between 5 to 10 hours spend 7 to 24 hours in sedentary and only 0 to 2 hours in very active mode. 
ggplot(data=merged_data, aes(x=total_minutes_asleep/60 ,y=sedentary_minutes/60, color=total_steps))+ 
  geom_point()+
  geom_smooth(method = 'loess', formula = 'y ~ x', color = '#005FCC') +
  scale_color_gradient(low="steelblue", high="orange") +
  labs(x = "Total Sleep Hours", y = "Sedentary Hours", title = "Sedentary Hours vs Sleep Hours")
  

ggplot(data=merged_data, aes(x=total_minutes_asleep/60 ,y=very_active_minutes/60, color=total_steps))+ 
  geom_point()+
  geom_smooth(method = 'loess', formula = 'y ~ x', color = '#005FCC') +
  scale_color_gradient(low="steelblue", high="orange")+
  labs(x = "total sleep hours", y = "very active hours")
 

ggplot(data=merged_data, aes(x=total_minutes_asleep, y = calories, color=total_minutes_asleep))+ 
  geom_point()+ 
  xlab("Total Minutes Alseep")+
  geom_smooth(method = 'loess', formula = 'y ~ x', color = '#005FCC') +
  scale_color_gradient(low="steelblue", high="orange")+
  labs(x = "total minutes asleep", y = "calories", title="Calories vs Total Minutes Asleep")

#calculate awake time
merged_data <- merged_data %>%
  mutate(awake_time = total_time_in_bed - total_minutes_asleep)

 ggplot(data = merged_data, aes(x = awake_time, y = sedentary_minutes, color = sedentary_minutes)) +
  geom_point() +
  scale_color_gradient(low="steelblue", high="orange")+
  geom_smooth(method = 'loess', formula = 'y ~ x', color = '#005FCC') +
  labs(title = "Sedentary minutes vs awake_time", x = 'awake_time', y = 'sedentary_minutes') +
  theme_classic()


plot13 <- ggplot(data = merged_data, aes(x = sedentary_minutes, y = total_minutes_asleep, color=total_minutes_asleep)) +
  geom_point() +
  scale_color_gradient(low="steelblue", high="orange")+
  geom_smooth(method = 'loess', formula = 'y ~ x', color = '#005FCC') +
  labs(title = "Total Minutes Asleep vs Sedentary minutes", x = 'sedentary minutes', y = 'minutes asleep') +
  theme_classic()
ggsave("total_minutes_asleep_vs_Sedentary_minutes.png", plot = plot13, width = 10, height = 6, dpi = 300) #plot13


```

















































































